---
output: pdf_document
fontsize: 11pt
geometry: margin=1in
---

```{r, include = FALSE }
knitr::opts_chunk$set( echo=FALSE, # hide code from document
                       warning=FALSE, # hide warnings 
                       message=FALSE) # hide messages 
library(tidyverse)
```

# STAA57 W21 Draft Report 

Group ? (member names)    

Link to RStudio Cloud shared project: https://rstudio.cloud/

***

## Introduction 

(*Description of questions that are being investigated*)

1. What are the trends in jet fuel costs over different time units (months/season)?
  - For this question we use the web scrapped to see how fuel price changes during time and we try to see how these changes affect the cost of flight or session for the company

2. What factors affect cost in a significant matter (model 1)
  - we decide here to use a linear regression model to assess significant facotrs to how cost is affected
  
3. What factors affect the duration of the same sets of exercises (model 2)
  - we decide here to use a linear regression model to assess significant facotrs to how duration of session is affected

4. What are the trends in demand over different time units (months/season/year)? 
 - we find how the demand in flight training changes and was season is more significant 

5. Considering the previous questions, what are the estimated average aircraft operating over different time units (months/season)?  

6. What are the total aircraft operating costs over month/season/year?  
  - assuming we know total number of planes
### Data 
```{r , results='hide', include=FALSE, warning=FALSE}
source(file = "preprocess_data.R")
```
(*Description of data used to answer the questions*)
For external data sources that we used in this project decided to gather from the following sources:

  - Fuel data: (https://www.indexmundi.com/commodities/?commodity=jet-fuel&months=240&currency=cad)
      - Historical fuel data by month, including the cost/gallon and %change
      - Assuming that all planes used in the flight school use the same fuel
  
  - Operating costs and repairs (https://cessna150152club.org/Costs):
      - The annual operating cost of a cessna plane model.
      - We made an assumption that there is a 10 dollar static cost added to each training session when a plane is used which we got from the source website
	
  - Weather (https://climate.weather.gc.ca/historical_data/search_historic_data_e.html):
		- Historical weather data by day, including temp, precipitation, wind, etc.
		- Gathered data from Oshawa location




## Analysis 

### Question 1: What are the trends in jet fuel costs over different time units (months/season)?
```{r}
fuel_by_month <- cleaned_fuel_prices %>%
  group_by(Month) %>%
  summarise(Avg_Price_Per_Gallon = mean(Price)) %>%
  ggplot(aes(x=Month, y=Avg_Price_Per_Gallon)) +
  geom_bar(stat="identity", fill="darkblue") +
  coord_cartesian(ylim=c(1.9, 2.2)) + 
  ylab("Avg Price Per Gallon") +
  labs(title="Fuel Price By Month ('16-'20)")

fuel_by_season <- cleaned_fuel_prices %>%
  group_by(Season) %>%
  summarise(Avg_Price_Per_Gallon = mean(Price)) %>%
  ggplot(aes(x=Season, y=Avg_Price_Per_Gallon)) +
  geom_bar(stat="identity", fill="darkblue") +
  coord_cartesian(ylim=c(1.9, 2.2)) + 
  ylab("Avg Price Per Gallon") +
  labs(title="Fuel Price By Season ('16-'20)")

ggarrange(fuel_by_month, fuel_by_season)
```
From the monthly data on the left, we can see that in March and April the fuel price is significantly cheaper than any other month at a price around \$1.91/gallon. Following that, it steadily increases and peaks around \$2.15/gallon around October and December.

For the seasonal data on the right, we can see that spring is significantly cheaper than any other season with a price at around \$1.94/gallon. On the other hand, fuel price is most expensive in the fall season with a price of about \$2.13/gallon

```{r, echo = FALSE, warning=FALSE, message=FALSE, fig.height=2}
clean_data_processed %>%
  distinct(Year, Month, Day, .keep_all = T) %>%
  mutate_if(is.numeric, replace_na, 0) %>%
  ggplot(aes(x=Date, y=`Fuel Price Per Gallon`)) + 
  geom_line(position="dodge", stat="identity", color="darkblue") +
  xlab('Time') + ylab('Fuel Price Per Gallon') +
  scale_x_date(breaks = 'year') +
  labs(title='Fuel Price Per Gallon Time Series')
```
From the above graph, it can be seen there was a significant decrease in the jet fuel price per gallon, leading to lower aircraft operating costs. Therefore, we expect that the average and total estimated aircraft costs to be lower in 2020.

### Question 3: What factors affect the duration of the same sets of exercises?
In order answer this, we first identified a set of weather factors that we believed
could have a strong effect on the duration of a session. These were:

* avg_relative_humidity
* avg_dew_point
* avg_pressure_sea
* avg_pressure_station
* avg_visibility
* avg_health_index
* precipitation
* avg_cloud_cover_4
* avg_temperature

Next, since we believed it was possible that the set of exercises chosen could directly affect the duration, we set it as a constant.
This was done by taking the top 3 most common exercises, and creating a model for each of them.
```{r}
library(broom)
data_exercises %>% 
  group_by(Exercises) %>%
  summarise(count=n()) %>%
  arrange(desc(count)) %>%
  slice(1:3) %>% 
  knitr::kable( caption = "Top 3 Freq. Exercises")

data_exercises_top_1 <- data_exercises %>%
  filter(Exercises == '16,17,18,30')

data_exercises_top_2 <- data_exercises %>%
  filter(Exercises == '16,17,18')

data_exercises_top_3 <- data_exercises %>%
  filter(Exercises == '16,17,18,23,30')
```

Here are the p-values for the factors from each of the 3 models:
```{r}
top_1_exercise_lm_out = lm(Total_Duration ~ 
                        avg_relative_humidity +
                        avg_dew_point +
                        avg_pressure_sea +
                        avg_pressure_station +
                        avg_visibility +
                        avg_health_index +
                        avg_cloud_cover_4 + 
                        avg_temperature + 
                        min_windchill +
                        precipitation + Aircraft - 1, 
                      data = data_exercises_top_1)
top_1_exercise_lm_out %>%
  tidy() %>%
  arrange(p.value) %>%
  select(term, p.value, estimate) %>% 
  knitr::kable( caption = "P values for '16,17,18,30'")

top_2_exercise_lm_out = lm(Total_Duration ~ 
                        avg_relative_humidity +
                        avg_dew_point +
                        avg_pressure_sea +
                        avg_pressure_station +
                        avg_visibility +
                        avg_health_index +
                        avg_cloud_cover_4 + 
                        avg_temperature + 
                        precipitation + Aircraft - 1, 
                      data = data_exercises_top_2)
top_2_exercise_lm_out %>%
  tidy() %>%
  arrange(p.value) %>%
  select(term, p.value, estimate) %>% 
  knitr::kable( caption = "P values for '16,17,18'")

top_3_exercise_lm_out = lm(Total_Duration ~ 
                        avg_relative_humidity +
                        avg_dew_point +
                        avg_pressure_sea +
                        avg_pressure_station +
                        avg_visibility +
                        avg_health_index +
                        avg_cloud_cover_4 + 
                        avg_temperature + 
                        precipitation + Aircraft - 1, 
                      data = data_exercises_top_3)
top_3_exercise_lm_out %>%
  tidy() %>%
  arrange(p.value) %>%
  select(term, p.value, estimate) %>% 
  knitr::kable( caption = "P values for '16,17,18,23,30'")
```
From these results we could not identify any common features between the three models.
Additionally, for all models, all factors have a p-value > 0.05 and thus individually are not statistically significant.
We can conclude that 

### Some Sub-section Title     


(*R output tables can be formatted nicely in .html with* `knitr::kable()`) 

```{r, echo = FALSE}
tibble( X = c("this", "my"), 
        Y = c("is", "table") ) %>% 
  knitr::kable( caption = "Table Caption")
```



## Summary 

(*Recap of your findings and conclusions*)


***  

# Appendix
## Import/Format Data 
```
library(tidyverse)
library(lubridate)
library(ggpubr)
library(rvest)
rm(list = ls())

raw_data = NULL
for( i in 1:17){
  tmp = readxl::read_xlsx( "data/UofT Data Set.xlsx", skip = 1, sheet = i, 
                           col_names =  paste( "X", 1:12, sep="" ) ) %>% 
    mutate( Instructor_ID = i,
            PPL = X1,
            X1 = replace( X1, !str_detect(X1, "Student"), NA ),
            PPL = zoo::na.locf( PPL ),
            X1 = zoo::na.locf(X1) )
  raw_data = bind_rows( raw_data, tmp )
}
rm(tmp,i)

names( raw_data  ) = c( "Student", "Year", "Month", "Day", "Aircraft", "LF_dual", 
                        "LF_solo", "Instrument_AC",  "Instrument_Sim", "CC_dual", "CC_solo", "Exercises",
                        "Instructor_ID", "Licence")

head(raw_data)

raw_data %>%
  filter( !is.na(Year), Year != "Year",
          Year >= 2016, Year <= 2020,) %>% 
  mutate_at( .vars = c(2:4), .funs = as.integer ) %>% 
  mutate_at( .vars = c(6:11), .funs = as.numeric ) %>% 
  mutate( Aircraft = str_to_upper(Aircraft),
          Aircraft = replace( Aircraft, str_detect(Aircraft, "GROUND"), "GROUND"),
          Aircraft = replace_na( Aircraft, "NA"),
          Aircraft = replace(Aircraft, Aircraft=="C152", "C-152"),
          Month = replace(Month, Month==111, 11),
          Other = ifelse( str_detect(Aircraft,"GROUND|NA"), -1, NA ),
          Student_ID = as.numeric( factor( paste( Student, Instructor_ID) ) ), 
          Session_ID = row_number() ) %>% 
  gather( key = "Training_Type", value = "Duration", 6:11, Other) %>% 
  filter( !is.na(Duration) ) %>% 
  mutate( Duration  = na_if(Duration, -1),
          Aircraft = na_if(Aircraft, "NA")) %>% 
  select( Instructor_ID, Student_ID, Session_ID, Year, Month, Day, 
          Aircraft, Duration, Training_Type, Exercises, Licence ) -> clean_data

getSeason <- function(month) {
  ifelse(month >= 3 & month <= 5, "Spring",
         ifelse(month >= 6 & month <= 8, "Summer",
                ifelse(month >= 9 & month <= 11, "Fall", "Winter")))
}

clean_data_processed = clean_data %>% 
  distinct( Session_ID, .keep_all = T) %>% 
  # split the exercises string into a "list" column w str_split()
  mutate( Exercises = str_split(Exercises, ",") ) %>%  
  # and expand list contents into multiple rows w/ unnest()
  unnest(Exercises) %>%
  # remove invalid exercises
  mutate(Exercises = as.integer(Exercises)) %>%
  filter(Exercises >= 1 & Exercises <= 30) %>%
  distinct_all() %>%
  mutate(
    Season = getSeason(Month), 
    Date = make_date(Year, Month, Day), 
    COVID = Year >= 2020
  )

# reading in the fuel per gallon price
webpage <- read_html("https://www.indexmundi.com/commodities/?commodity=jet-fuel&months=240&currency=cad")
tbls <- html_nodes(webpage, "table") %>%
  html_table(fill = TRUE)
fuel_prices <- as.data.frame(tbls[2])
colnames(fuel_prices) <- c("Month_Year", "Price","Change")
clean_data_processed$Month_Words <- month.abb[clean_data_processed$Month]
clean_data_processed$Month_Year <- (str_c(clean_data_processed$Month_Words, clean_data_processed$Year, sep = " "))

clean_data_processed <- left_join(clean_data_processed,fuel_prices,by="Month_Year",copy = TRUE)
clean_data_processed <- select(clean_data_processed, -c("Month_Year","Change","Month_Words"))
colnames(clean_data_processed)[colnames(clean_data_processed) == 'Price'] <- 'Fuel Price Per Gallon'

process_weather <- function(data) {
  data %>% 
    select(
      'Date/Time', 
      "Mean Temp (°C)",
      "Total Precip (mm)",
      "Total Rain (mm)", 
      "Total Snow (cm)", 
      "Spd of Max Gust (km/h)"
    ) %>% 
    mutate(wind = as.character(`Spd of Max Gust (km/h)`)) %>%
    select(-c("Spd of Max Gust (km/h)")) %>%
    mutate(wind = ifelse(wind == '<31', 31, wind))
}

climate_data_2016 <- read_csv('data/en_climate_daily_ON_6158410_2016_P1D.csv') %>% process_weather()
climate_data_2017 <- read_csv('data/en_climate_daily_ON_6158410_2017_P1D.csv') %>% process_weather()
climate_data_2018 <- read_csv('data/en_climate_daily_ON_6158410_2018_P1D.csv') %>% process_weather()
climate_data_2019 <- read_csv('data/en_climate_daily_ON_6158410_2019_P1D.csv') %>% process_weather()
climate_data_2020 <- read_csv('data/en_climate_daily_ON_6158410_2020_P1D.csv') %>% process_weather()
climate_data <- bind_rows(climate_data_2016, 
                          climate_data_2017, 
                          climate_data_2018, 
                          climate_data_2019, 
                          climate_data_2020)

clean_data_processed <- left_join(clean_data_processed, climate_data, by=c("Date" = "Date/Time"), copy=True)

```



