---
output: pdf_document
fontsize: 11pt
geometry: margin=1in
---

```{r, include = FALSE }
knitr::opts_chunk$set( echo=FALSE, # hide code from document
                       warning=FALSE, # hide warnings 
                       message=FALSE) # hide messages 
library(lubridate)
library(modelr)
library(stringi)
library(ggpubr)
library(rvest)
library(broom)
library(MASS)
library(tidyverse)
library(fastDummies)
```

# STAA57 W21 Draft Report 

Group 4 (Adham F, Jason Y, Mohamed T, Wesley M)

Link to RStudio Cloud shared project: https://rstudio.cloud/spaces/115177/project/2202760

***

## Introduction 

The topic of this paper is to investigate the aircraft operating cost of DFC, particularly the Cost per Session. There are two main factors that affect the aircraft operating cost which are fuel price (we are using gasoline) and the duration of the session. It is intuitive to think that the duration of a flight will be affected by the weather. Factors such as cloud coverage, wind and precipitation can affect an aircraft/pilot’s performance, thus the duration of flight/lesson. With an increase in flight duration comes an increase in maintenance and fuel costs. Therefore, we should expect that there will be weather factors in our data that can be significant predictors for the aircraft operating cost to a certain extent. It is also known that fuel price does change monthly/seasonally depending on seasonal travel, such as during Christmas and Summer. Therefore, we should expect that there are seasonal changes to the aircraft operating costs. Since seasonal changes in fuel costs are fairly obvious and weather data is closely correlated to seasonal weather changes, we expect the weather factors that we are using to be able to predict the aircraft operating cost. In addition, since COVID-19 has dramatically impacted the Airline industry, which is closely related to fuel, we will also be investigating COVID-19 being a factor in predicting the aircraft operating costs.

### Questions:
1. What are the trends in fuel costs over different time units (daily/seasonly)?
    - For this question, web scrapping was used to see how fuel price changes over time and attempted to see how these changes affect the Cost per Session for DFC.
    - A monthly analysis was considered, however due to the time period, it would result in a relatively hard to interpret analysis of the fuel price fluctuations that can also be deduced using a seasonal time period analysis.

For the next two questions we will use Linear Regression models. For context:

This is a method to predict numerical values like cost/duration given new input data (exercise plan, season, weather, fuel, etc.). The model is essentially a math equation generated using the given historical data plus external factors we add (e.g. weather, fuel cost).

From this model, we can also see which factors have a significant effect on what we are trying to predict.

2. What factors affect the duration of the sessions?
    - We will use a linear regression model to assess the significant factors associated with the duration of the sessions.

3. What factors affect cost in a significant matter?
    - We will repeat the same process as the previous question, however predicting cost instead of duration.

## Data 
```{r , results='hide', include=FALSE, warning=FALSE}
rm(list = ls())

raw_data = NULL
for( i in 1:17){
  tmp = readxl::read_xlsx( "data/UofT Data Set.xlsx", skip = 1, sheet = i, 
                           col_names =  paste( "X", 1:12, sep="" ) ) %>% 
    mutate( Instructor_ID = i,
            PPL = X1,
            X1 = replace( X1, !str_detect(X1, "Student"), NA ),
            PPL = zoo::na.locf( PPL ),
            X1 = zoo::na.locf(X1) )
  raw_data = bind_rows( raw_data, tmp )
}
rm(tmp,i)

names( raw_data  ) = c( "Student", "Year", "Month", "Day", "Aircraft", "LF_dual", 
                        "LF_solo", "Instrument_AC",  "Instrument_Sim", "CC_dual", "CC_solo", "Exercises",
                        "Instructor_ID", "Licence")

head(raw_data)

raw_data %>%
  filter( !is.na(Year), Year != "Year",
          Year >= 2016, Year <= 2020,) %>% 
  mutate_at( .vars = c(2:4), .funs = as.integer ) %>% 
  mutate_at( .vars = c(6:11), .funs = as.numeric ) %>% 
  mutate( Aircraft = str_to_upper(Aircraft),
          Aircraft = replace( Aircraft, str_detect(Aircraft, "GROUND"), "GROUND"),
          Aircraft = replace_na( Aircraft, "NA"),
          Aircraft = replace(Aircraft, Aircraft=="C152", "C-152"),
          Month = replace(Month, Month==111, 11),
          Other = ifelse( str_detect(Aircraft,"GROUND|NA"), -1, NA ),
          Student_ID = as.numeric( factor( paste( Student, Instructor_ID) ) ), 
          Session_ID = row_number() ) %>% 
  gather( key = "Training_Type", value = "Duration", 6:11, Other) %>% 
  filter( !is.na(Duration) ) %>% 
  mutate( Duration  = na_if(Duration, -1),
          Aircraft = na_if(Aircraft, "NA")) %>% 
  select( Instructor_ID, Student_ID, Session_ID, Year, Month, Day, 
          Aircraft, Duration, Training_Type, Exercises, Licence ) -> clean_data

# filter and clean exercises
clean_data <- clean_data %>%
  filter(Aircraft != "GROUND") %>%
  filter(!is.na(Exercises)) %>%
  # not sure what exercise 40 is, so we will remove it
  filter(!str_detect(Exercises, "40")) %>%
  # fix typos
  mutate(Exercises = str_replace(Exercises, "`7", "17")) %>%
  mutate(Exercises = str_replace(Exercises, "112", "12")) %>%
  mutate(Exercises = str_replace(Exercises, "81", "18")) %>%
  mutate(Exercises = str_replace(Exercises, "55", "5")) %>%
  mutate(Exercises = str_replace(Exercises, "2.4", "2,4")) %>%
  mutate(Exercises = str_replace(Exercises, "16 17", "16,17")) %>%
  mutate(Exercises = str_replace(Exercises, "67", "6,7")) %>%
  mutate(Exercises = str_replace(Exercises, "1821", "18,21")) %>%
  mutate(Exercises = str_replace(Exercises, "1314", "13,14")) %>%
  mutate(Exercises = str_replace(Exercises, "1316", "13,16")) %>%
  mutate(Exercises = str_replace(Exercises, "1617", "16,17")) %>%
  mutate(Exercises = str_replace(Exercises, "18.30", "18,30")) %>%
  mutate(Exercises = str_replace(Exercises, "34", "3,4")) %>%
  mutate(Exercises = str_replace(Exercises, "89", "8,9")) %>%
  # remove trailing comma
  mutate(Exercises = str_replace(Exercises, ",$", "")) %>% 
  # remove leading comma
  mutate(Exercises = str_replace(Exercises, "^,", "")) %>%
  # remove duplicate comma
  mutate(Exercises = str_replace(Exercises, ",,", ","))

getSeason <- function(month) {
  ifelse(month >= 3 & month <= 5, "Spring",
         ifelse(month >= 6 & month <= 8, "Summer",
                ifelse(month >= 9 & month <= 11, "Fall", "Winter")))
}

clean_data_processed = clean_data %>% 
  distinct( Session_ID, .keep_all = T) %>% 
  # split the exercises string into a "list" column w str_split()
  mutate( Exercises = str_split(Exercises, ",")) %>%  
  # and expand list contents into multiple rows w/ unnest()
  unnest(Exercises) %>%
  # remove invalid exercises
  mutate(Exercises = as.integer(Exercises)) %>%
  filter(Exercises >= 1 & Exercises <= 30) %>%
  group_by(Session_ID) %>%
  distinct_all() %>%
  mutate(
    Season = getSeason(Month), 
    Date = make_date(Year, Month, Day), 
    COVID = Year >= 2020
  )

gasoline_price <-read_csv("data/18100001.csv")
fuel_prices<-gasoline_price%>%filter(GEO=="Toronto, Ontario")%>%filter(str_detect(REF_DATE,"(2016|2017|2018|2019|2020|2021)-\\d{2}"))%>%group_by(REF_DATE)%>%summarise(avg_price=mean(VALUE)*3.78541/100)%>%separate(REF_DATE, c("Year", "Month"), "-")%>%mutate(Month = month.abb[as.numeric(Month)])%>%mutate(Month_Year=str_c(Month,Year,sep=" "))%>%mutate(Price=avg_price)%>%select(-c("Month","Year","avg_price"))

cleaned_fuel_prices <- fuel_prices %>%
  separate(Month_Year, c("Month", "Year"), " ") %>%
  filter(Year >= 2016, Year <= 2020) %>%
  mutate(Month = factor(Month, level=month.abb)) %>%
  mutate(Season = getSeason(match(Month, month.abb))) %>%
  mutate(Season = factor(Season, level=c("Spring", "Summer", "Fall", "Winter")))

clean_data_processed$Month_Words <- month.abb[clean_data_processed$Month]
clean_data_processed$Month_Year <- (str_c(clean_data_processed$Month_Words, clean_data_processed$Year, sep = " "))

clean_data_processed <- left_join(clean_data_processed,fuel_prices,by="Month_Year",copy = TRUE)
clean_data_processed <- select(clean_data_processed, -c("Month_Year","Month_Words"))
colnames(clean_data_processed)[colnames(clean_data_processed) == 'Price'] <- 'Fuel Price Per Gallon'

process_weather <- function(data) {
  data %>% 
    select(
      "date",
      "avg_relative_humidity",
      "avg_hourly_wind_speed",
      "avg_dew_point",
      "avg_pressure_sea",
      "avg_pressure_station",
      "avg_visibility",
      "avg_health_index",
      "precipitation",
      "avg_cloud_cover_4",
      "avg_temperature"
    ) %>%
    filter(date <"2021-01-01", date >= "2016-01-01")
}

climate_data <- read_csv('data/weatherstats_oshawa_daily.csv') %>% process_weather()

# ----- BEGIN NEW CODE -----
data_exercises <- clean_data %>% 
  group_by(Session_ID, Exercises, Year, Month, Day, Aircraft) %>% 
  # can have multiple training types, therefore get total duration per session
  summarise(Total_Duration = sum(Duration, na.rm = T)) %>%
  ungroup() %>%
  mutate(
    N_Exercises = stri_count(Exercises, fixed = ",")+1,
    Season = getSeason(Month), 
    Date = make_date(Year, Month, Day),
    Month_Words = month.abb[Month]
  ) %>% 
  mutate(Month_Year = (str_c(Month_Words, Year, sep = " "))) %>%
  left_join(climate_data, by=c("Date" = "date")) %>%
  left_join(fuel_prices, by="Month_Year") %>%
  mutate(`Fuel Price Per Gallon` = Price) %>%
  mutate(COVID = Year >= 2020) %>%
  # add dummy variables for all of the exercises
  dummy_cols(split=',', select_columns=c("Exercises")) %>%
  select(-c(Month_Year, Month_Words, Price))
# ----- END NEW CODE -----

clean_data_processed <- left_join(clean_data_processed, climate_data, by=c("Date" = "date"), copy=True)
```

The external data that is used in this project were gathered from the following sources:
  
  - Fuel (https://open.canada.ca/data/en/dataset/cbb9182b-df94-4ae6-8a85-b75a34b2f990)
      - Historical fuel data by month, including the cost/gallon
      - Assuming that all planes used in the flight school use the same fuel
      - From preliminary research, it was found that C-172, C-150 and C-152 aircraft models only consume AVGas and gasoline. We decided to use gasoline data for our analysis. 
  - Maintenance costs and repairs (https://cessna150152club.org/Costs):
      - The annual operating cost of a Cessna plane model for maintenance and repairs
      - Based on this research, an assumed 15 dollars static cost per hour was added to each training session cost
      - In addition, the fuel consumption per hour is estimated to be 6 gallons
  - Weather (https://oshawa.weatherstats.ca/download.html):
      - Historical weather data by day, including temp, precipitation, wind, etc.
      - Gathered data from Oshawa location

## Analysis 
### Question 1: What are the trends in fuel costs over different time units (daily/season)?

```{r, echo = FALSE, warning=FALSE, message=FALSE, fig.height=2}
clean_data_processed %>%
  distinct(Year, Month, Day, .keep_all = T) %>%
  mutate_if(is.numeric, replace_na, 0) %>%
  ggplot(aes(x=Date, y=`Fuel Price Per Gallon`)) + 
  geom_line(position="dodge", stat="identity", color="darkblue") +
  xlab('Time') + ylab('Fuel Price Per Gallon') +
  scale_x_date(breaks = 'year') +
  labs(title='Fuel Price Per Gallon Time Series')
```
From the above graph, it can be seen there was a drop in the fuel price per gallon in 2020, leading to lower aircraft operating costs. Therefore, we expect that the Cost per Session to be lower during 2020, i.e. during the pandemic.

```{r, fig.height=2}
fuel_by_season <- cleaned_fuel_prices %>%
  group_by(Season, Year) %>%
  summarise(Avg_Price_Per_Gallon = mean(Price)) %>%
  ggplot(aes(x=Season, y=Avg_Price_Per_Gallon, fill=Year)) +
  geom_bar(stat="identity", position="dodge") +
  ylab("Avg Price Per Gallon") +
  labs(title="Fuel Price By Season ('16-'20)")
fuel_by_season
```

The above graph shows the average price per season from 2016 to 2020. There is no clear pattern to the fluctuations of the gasoline prices over the season. It is interesting to note that the gasoline prices during the pandemic are not at its lowest as for almost all the seasons 2016 had the lowest average price per gallon, therefore that is an indicator that the pandemic may not be an important factor when analysing the Cost per Session model.

### Question 2: What factors affect the duration of sessions?

**How was the data prepared?**

For this model, we combined the given flight data with Oshawa weather data.
In particular, we referenced https://silicium.dk/fsx/FSWeb/LearningCenter/TheWorld/WeatherAviationWeatherBasics.htm, 
and identified a set of weather factors that were deemed to have a strong effect on the duration of a session.

* `Avg relative humidity`
* `Avg dew point`
* `Avg sea pressure`
* `Avg visibility`
* `Avg health index` (air quality)
* `Precipitation`
* `Avg cloud cover `
* `Avg temperature`

In total, our factors consist of the selected weather factors, the exercises conducted, the season, and whether or not the session took place in 2020 (Covid year).

We also spit our data into two sets at an 8:2 ratio:

  1. Training (80% of data) set is used to build the equation and determine the significant factors.
  2. Test (20% of data) set is used to test accuracy of model.


Linear regression model for Duration per Session:

```{r}
set.seed(123)
duration.train <- data_exercises %>% sample_frac(.80)
duration.test  <- data_exercises %>% setdiff(duration.train)
# Training LM model
duration.fit = lm(Total_Duration ~ 
                        avg_dew_point +
                        avg_pressure_sea +
                        avg_visibility +
                        avg_health_index +
                        avg_cloud_cover_4 + 
                        avg_temperature + 
                        COVID +
                        Exercises_1 +
                        Exercises_2 +
                        Exercises_3 +
                        Exercises_4 +
                        Exercises_5 +
                        Exercises_6 +
                        Exercises_7 +
                        Exercises_8 +
                        Exercises_9 +
                        Exercises_10 +
                        Exercises_11 +
                        Exercises_12 +
                        Exercises_13 +
                        Exercises_14 +
                        Exercises_15 +
                        Exercises_16 +
                        Exercises_17 +
                        Exercises_18 +
                        Exercises_19 +
                        Exercises_20 +
                        Exercises_21 +
                        Exercises_22 +
                        Exercises_23 +
                        Exercises_24 +
                        Exercises_29 +
                        Exercises_30 +
                        precipitation + Season - 1, 
                      data = duration.train)

tidy(duration.fit) %>%
  filter(p.value < 0.05) %>%
  mutate(minutes = 60*estimate) %>%
  select(term, minutes) %>%
  knitr::kable( caption = "Most Significant Factors")

duration.test %>% 
  add_predictions(duration.fit, var="pred") %>% 
  drop_na(Total_Duration, pred) %>%
  summarise(minutes = 60 * sum(abs(`Total_Duration` - `pred`))/n()) %>%
  knitr::kable( caption = "MAE") ->
  duration.mae
```

**Analysis for Duration per Session linear regression model**

From our model, we can see that the most significant factors for predicting duration are:

* Exercises_12: Stall
* Exercises_16: Take-Off
* Exercises_19: First Solo
* Exercises_23: Pilot Navigation
* Exercises_24: Instrument Flying
* Exercises_30: Radio Communications
* precipitation: How much avg precipitation received that day
* SeasonSummer: Whether or not the session took place in the summer

The model predicts that on average:

* Having Exercise 19 (first solo) decreases the duration by 44min
* Every mm of precipitation decreases the duration by 0.4min
* The session being in the summer increases the duration by 4.7min
* etc.
For the remaining factors refer to *Figure 1* in the appendix.

Our model has an R^2 value of 0.88, meaning that about 88% of the changes in the duration can be explained by changes in our factors.
To test the model's accuracy, we tried predicting the duration by inputting the test data, and compared the results to the actual duration of the test sessions. We then took the average of the differences (Mean Absolute Error or MAE) and got an average error of ~14min.
This means that using our model to predict new data will result in a value that's off by 14min on average.

### Question 3: What factors affect cost in a significant matter?
**How was the data prepared?**

The aircraft operating cost per session was calculated using the equation below:

`CostPerSession = (Duration * (FuelPricePerGallon*GallonsPerHr + OperatingCostPerHr))`

The aircraft operating cost equation uses duration, fuel price per gallon, and an average number of gallons consumed per hour (6) as well as an estimated maintenance and repairs cost per hour (15). After the Cost per Session column has been calculated, the data frame is then grouped by date and an Average Cost per Session is calculated for each date to smoothly join the data frame with the daily weather data. A linear regression model can then be built using this data frame.

Since the cost of each session is dependent on the duration, it was observed that session duration can vary due to the exercises being performed as well as the number of exercises. Some sessions might have more exercises and thus have a higher cost. Due to the lack of information on the expected individual exercise duration, it was assumed that each exercise will have the same duration. Since the cost of a session is related to the exercises being performed, we added a column representing the number of exercises in a session and added that to our model.

```{r}
# NOTE that this is using the same data as was used for duration. It is showing
# cost per session rather than daily avg of cost per session.
data_exercises %>% 
  group_by(Session_ID) %>%
  mutate(`Cost Per Session`=(Total_Duration*`Fuel Price Per Gallon`*6 + 15*Total_Duration)) %>%
  #distinct(Session_ID, .keep_all = T) %>%
  mutate(`Average Cost Per Session`= mean(`Cost Per Session`)) %>%
  arrange(Date) %>%
  tibble -> cost_per_session

# Training data
set.seed(123)
cost.train <- cost_per_session %>% sample_frac(.8)
cost.test <- cost_per_session %>% setdiff(cost.train)
cost.fit = lm(`Cost Per Session` ~ 
                        avg_dew_point +
                        avg_pressure_sea +
                        avg_visibility +
                        avg_health_index +
                        avg_cloud_cover_4 + 
                        avg_temperature +
                        COVID +
                        Exercises_1 +
                        Exercises_2 +
                        Exercises_3 +
                        Exercises_4 +
                        Exercises_5 +
                        Exercises_6 +
                        Exercises_7 +
                        Exercises_8 +
                        Exercises_9 +
                        Exercises_10 +
                        Exercises_11 +
                        Exercises_12 +
                        Exercises_13 +
                        Exercises_14 +
                        Exercises_15 +
                        Exercises_16 +
                        Exercises_17 +
                        Exercises_18 +
                        Exercises_19 +
                        Exercises_20 +
                        Exercises_21 +
                        Exercises_22 +
                        Exercises_23 +
                        Exercises_24 +
                        Exercises_29 +
                        Exercises_30 +
                        precipitation + Season - 1, 
                      data = cost.train)
summary(cost.fit)

glance(cost.fit) %>%
  knitr::kable( caption = "Model Results")

tidy(cost.fit) %>%
  filter(p.value < 0.05) %>%
  knitr::kable( caption = "Most Significant Factors")

# Testing LM model
cost.test %>% 
  add_predictions(cost.fit, var="pred") %>%
  drop_na(`Cost Per Session`, pred) %>%
  summarise(mae = sum(abs(`Cost Per Session` - `pred`))/n()) %>%
  knitr::kable( caption = "MAE")
# clean_data_processed %>% 
#   group_by(Session_ID) %>%
#   mutate(`Cost Per Session`=(Duration*`Fuel Price Per Gallon`*6 + 15*Duration), n_exercises=n()) %>%
#   distinct(Session_ID, .keep_all = T) %>%
#   group_by(Date) %>%
#   mutate(`Average Cost Per Session`= mean(`Cost Per Session`)) %>%
#   arrange(Date) -> daily_average_cost_per_hour
# 
# fit.full = lm(`Average Cost Per Session` ~ `avg_temperature` + `avg_relative_humidity`+ `avg_dew_point` + `avg_pressure_sea` + `avg_pressure_station` + `avg_visibility` + `avg_health_index` + `precipitation` + `avg_cloud_cover_4`+ COVID + Season + n_exercises - 1, data=daily_average_cost_per_hour)
```

### Interpreting the coefficients statistically significant factors

The analysis below for the COVID-19 factors assumes that every other variable is held constant:

* `SeasonSummer` - during summer the cost increases by 1 dollar per session

* `SeasonWinter` - during winter the cost increases by 2.26 dollars per session

* `avg_health_index` - for each unit increase in the `avg_health_index` the cost increases by 0.74 dollars per session

* `n_exercises` - for each additional exercise the cost decreases by 0.13 dollars per session

* `avg_pressure_sea` - for each unit increase in the `avg_pressure_sea` the cost decreases by 31 dollars per session

* `avg_pressure_station` - for each unit increase in the `avg_pressure_station` the cost increases by 32 dollars per session

* `precipitation` - for each unit increase in the precipitation the cost decreases by 0.07 dollars per session

* `avg_cloud_cover_4` - for each unit increase in the precipitation the cost decreases by 0.34 dollars per session

An interesting observation is that as the number of exercises increasing is associated with the cost decreasing - this does not mean causation, rather just association, meaning that increasing the number of exercises doesn't result in a decrease in cost. A possible explanation is that there can be a lot of easy exercises for beginner sessions that don't take long to complete therefore lowering the duration of the session thus lowering the cost per session.

Unlike what we anticipated from question 1 that Fall would have higher cost, it turned out to be lower than the usual cost as the coefficient for the fall was used as the base state for the equation and the other seasons such as Summer and Winter had a statistically significant coefficients. It was during the Winter that the price per session increases by 2.26 dollars as anticipated to have a spike around December in the fuel cost analysis. Unintuitively, this model shows that COVID-19 had no effect on the price per session due to its high p-value. However, it is important to keep in mind that association  does not mean causation, therefore this does not rule out that COVID-19 cannot affect the price per session.

## Summary 

Overall, the Duration per Session model was not as useful as anticipated due to linear regression not being a good fit for the data, and that can be the result of choosing the incorrect model type. Nonetheless, the second model did indeed provide some useful insight on how the Cost per Session changes depending on the statistically significant factors, despite going against some anticipated patterns found using the fuel analysis. It is important to note that the residual plot for the Cost per Session did not appear to follow a random pattern thus there can be a better model for the data.

***  

# Appendix
*Figure 1*
```{r}
tidy(duration.fit) %>%
  mutate(minutes = 60*estimate) %>%
  select(term, minutes) %>%
  knitr::kable( caption = "Most Significant Factors")
```
